\begin{center}
    \large\textsc{Acknowledgements}
\end{center}
\markboth{Acknowledgements}{}
\addcontentsline{toc}{section}{Acknowledgements}

This thesis would not have been possible without the supervision, support, kindness and love I received from so many people througout the years. I will certainly not be able to name everyone, so I want to start off with a big \textit{thank you} to everyone who made the last years such a fun, exciting, educational and rewarding experience. \\

First, my deepest gratitude belongs to Barbara, for giving me the opportunity to work on exciting topics with a bunch of amazing people. Your awe-inspiring compentence and admirable kindness made every discussion with you an uplifting and insightful experience. A huge thanks also goes to Gray, your vision and passion for science sparks enthusiasm and reignited my motivation many times. The two of you have assembled a team of excellent scientists and created a space buzzing with inovation. It was just a blast to work with every single one in the TreutCamp ensemble, thanks to all of you! Particularly I want to express my gratitude towards Fátima, Sophie and Fides, the shared first authors of the studies presented here. Quite literally none of this would have been possible without your know-how and talent. I'll be forever grateful for your (sometimes also emotional) support. Moreover, a great thanks to Zhisong for beeing a guiding light especially in the first years of my PhD. I learned quite a bit of new skills over the years and I feel like I owe a significant portion of this to your deep knowledge and experience. \\

I want to thank our external collaborators and co-authors from Vienna, Prof. Jürgen Knoblich and Dr. Chong Li, who gave me the phenomenal chance to expand my horizon and work on such an exciting project. I also want to thank Prof. Fabian Theis and his group, for the invitation to Munich and many fruitful and enlightening discussions. \\

Additionally, I want to thank Prof. Fabian Theis along with Prof. Randall Platt, who kindly agreed to participate in my thesis examination committee and Prof. Sai Reddy for chairing the defense. \\

I want to thank Dr. Georg Zeller, my master thesis supervisor for his support even during my PhD and for lastingly shaping my view on science. \\

Science really wouldn't be as fun without all the fascinating scientific discussions with incredibly smart people. I therefore want to thank everyone who I had the pleasure of exchanging interesting ideas with, including (but certainly not limited to) Christoph Harmel, Dr. Bastian Rieck, Dr. Tomás Gomes, Gilles Gut, Dr. Giovanna Brancati, Dr. Qianhui Yu, Dr. Matteo Togninally, Dr. Doris Popovic, Dr. Akanksha Jain and Dr. Philipp Wahle. \\

This also feels like the right place to thank Dr. Makiko Seimiya, Dr. Malgorzata Santel and Dr. Ryoko Okamoto for keeping the lights on in the wetlab (even though I wasn't really there a lot). \\

I am also deeply grateful for all friends who made the last years worthwile. First and foremost, I have to thank Simon, Benny and Chris for beeing the best and closest long-distance friends one could ever wish for. Thank you for everything! For prophecies, conspiracies, new holidays, nights, days and simply for existing. I want to thank  Christoph, for walks and talks and herbs. Also Lisi, Jamie, Jana, the Brits, the Cocktail Crew, all climbing and bouldering buddies and everyone else who made my time in Basel so enjoyable. Thanks also to Phil, Nico, Jana, Lasse, Tob, Jyoti, Lizzy, Jutta, Jannik, Pascal, and everyone else who are sadly not in Basel (but mostly in Berlin). \\

I'm also incredibly thankful for Juliane, who stuck with me throughout the years. You enrich my life in so many ways and you in bring out the best in me. You are amazing, and I love you. \\

Zu guter Letzt möchte ich meiner Familie danken: Meiner Mutter Karin, meinem Bruder Sammy, meinen Großeltern Gerda, Horst, Helga und Ernst, meinem Stiefvater Stephan und meinen Stiefgeschwistern Jonas, Karen, Philipp und Anna. Eure unerschütterliche Unterstützung und Fürsorge hat micht durch mein Leben getragen und mich zu der Person gemacht, die ich heute bin. Ich hab euch unglaublich lieb. 



